---
params: 
  location: "Montana"
  asset_dir: "./assets"
  attr_id: "Division"
  attr_name: "Climate Division"
  attr_plural: "Climate Divisions"

execute: 
  cache: true
---

```{r setup, echo=FALSE}
library(magrittr)
library(ggplot2)

shp <- sf::read_sf(file.path(params$asset_dir, "agg_regions.geojson"))
mt <- normals::mt %>% 
  dplyr::mutate(state = "MT")

f_list <- RCurl::getURL(
  "https://data.climate.umt.edu/mca/cmip/",
  ftp.use.epsv = TRUE,
  dirlistonly = FALSE
) %>%
  XML::getHTMLLinks() %>%
  grep(".tif", ., value = T) %>%
  grep(".json", ., value = T, invert = T)

cmip_files_full <- file.path(
  "https://data.climate.umt.edu/mca/cmip",
  basename(f_list)
)

cmip_files <- list.files(params$asset_dir, pattern = ".rds", full.names = T)

climdiv_to_factor <- function(x) {
  factor(x, levels = c("Western", "Southwestern", "North Central", "Central", "South Central", "Northeastern", "Southeastern"))
}

tas_tsfm <- function(x) (x - 273.15) * 1.8 + 32
pr_tsfm <- function(x) x/25.4

inc_or_dec <- function(x) {
  x %<>% round(5)
  if (x < 0) {
    return("decrease")
  } else if (x > 0) {
    return("increase")
  } else {
    return("stay the")
  }
}

agreements <- readr::read_csv(
  "./assets/agreements.csv",
  show_col_types = FALSE
) %>% 
  dplyr::mutate(avg = round(avg, 2), agree = round(agree))

most_common <- function(x) {
  names(sort(table(x), decreasing = T)[1])
}

subset_agreements <- function(variable, period, scenario=NULL, fun=NULL, col=NULL) {
  library(magrittr)
  tmp <- agreements %>% 
    dplyr::filter(
      variable == !!variable,
      period == !!period
    ) 
    
  if (!is.null(scenario)) {
    tmp %<>% 
      dplyr::filter(scenario == !!scenario)
  }
  
  if (!is.null(col)) {
    tmp %<>% dplyr::pull(col)
  }
  
  if (!is.null(fun)) {
    tmp %<>% fun()
  }
  
  return(tmp)
}

make_bullet_text <- function(period, variable, suffix, description) {
  if (period == "mid") {
    long = "Mid-century"
    short = "mid_century"
  } else {
    long = "End-of-century"
    short = "end_century"
  }
  up_down_txt = subset_agreements(variable, short, fun=most_common, col='txt')
  agree <- subset_agreements(variable, short)
  mn_num <- min(agree$avg)
  mn_agree <- dplyr::slice(agree, which.min(avg)) %$% agree
  
  mx_num <- max(agree$avg)
  mx_agree <- dplyr::slice(agree, which.max(avg)) %$% agree
  glue::glue("- {long}: Across the domain, the majority of SSP scenarios project that the {description} in the {tolower(long)} will {up_down_txt}. Changes in {description} across SSPs range from {mn_num} {suffix} ({mn_agree}% model agreement) to {mx_num} {suffix} ({mx_agree}% model agreement)")
}
```

# **Draft**  Future Projections
## Global Climate Modeling
Projecting future climate on a global scale requires modeling many intricate relationships between the land, ocean, and atmosphere. Many global climate and Earth system models exist, each varying in complexity, capabilities, and limitations.

Consider one of the simplest forms of a model used for future projections, a linear regression model (@fig-lm-example). With this model, researchers would plot a climate variable (e.g., temperature) over time, draw a best-fit, straight line through the data, and then extend the line into the future. That line, then, provides a means of projecting future conditions. Whether or not those projections are valid is a separate question. For example, the model may be based on false assumptions: the relationship may a) not be constant through time, b) not include outside influences such as human interventions (e.g., policy regulations), and c) not consider system feedbacks that might enhance or dampen the relationship being modeled.

```{r echo=FALSE, warning=FALSE, message=FALSE}
#| label: fig-lm-example
#| fig-cap: "Example of fitting a linear model to data to predict future conditions."
x = 1981:2045
y = seq(30, 35, length.out = length(x)) + rnorm(length(x), 0, 1)
z = seq(30, 35, length.out = length(x))

tibble::tibble(x = x, y = y) %>% 
  dplyr::mutate(
    future = ifelse(x > 2022, "Projected", "Observed"),
    y = ifelse(future == "Projected", NA, y)
  ) %>% 
  ggplot() + 
    geom_point(aes(x=x, y=y), color="#F8766D") +
    geom_smooth(aes(x=x, y=z, linetype=future), method = "lm", se = FALSE, color="black") + 
    theme_minimal() + 
  geom_vline(aes(xintercept=2022)) + 
  labs(x="Year", y="Average Temperature (°F)", color="", linetype = "")
```

While the linear regression model provides an instructive visual aid for considering modeling, it is too simple for looking at climate changes, in which the interactions are complex and often nonlinear. For example, if temperatures rises, evaporation is expected to increase. At the same time, increasing temperatures increase the atmosphere’s capacity to hold water. Water is a greenhouse gas so more water in the atmosphere means the atmosphere can absorb more heat... thus driving more evaporation. What seemed a simple relationship has changed (possibly dramatically) because of this feedback between temperature, evaporation, and the water-holding capacity of the atmosphere. 

Linear models do not account for such nonlinear relationships. Instead, climate scientists account for nonlinearity through computer simulations that describe the physical and chemical interactions between the land, oceans, and atmosphere. These simulations, which project climate change into the future, are called general circulation models (GCMs; see sidebar)

::: {.callout-note icon=false}
### General Circulation Models {#gcm-callout}

General circulation models (GCMs) help us project future climate conditions. They are the most advanced tools currently available for simulating the response of the global climate system—including processes in the atmosphere, ocean, cryosphere, and land surface—to increasing greenhouse gas concentrations. 

GCMs depict the climate using a 3-D grid over the globe, typically having a horizontal resolution of between 250 and 600 km (160 and 370 miles), 10-20 vertical layers in the atmosphere and sometimes as many as 30 layers in the oceans. Their resolution is quite coarse. Thus, impacts at the scale of a region, for example for Montana, require downscaling the results from the global model to a finer spatial grid (discussed later) (text adapted from IPCC 2013b).
:::

Because of the complexities involved, climate scientists rarely rely on a single model, but instead use an ensemble (or suite) of models. Each model in an ensemble represents a single description of future climate based on specific initial conditions and assumptions. The use of multiple models helps scientists explore the variability of future projections (i.e., how certain are we about the projection) and incorporate the strengths, as well as uncertainties, of multiple approaches.

For the work of the Montana Climate Assessment, we employed an ensemble from the sixth iteration of the Coupled Model Intercomparison Project (CMIP6), which includes over 100 GCMs depending on the experiment conducted (@cmip6). The World Climate Research Program describes CMIP as “a standard experimental protocol for studying the output” of GCMs (@usgcrp). It provides a means of validating, comparing, documenting, and accessing diverse climate model results. The CMIP project dates back to 1995, with the sixth iteration (CMIP6) starting in 2016 and providing climate data for the latest IPCC Sixth Assessment Report (@ipcc).

We employed 8 individual GCMs from the CMIP6 project for the Montana Climate Assessment ensemble, chosen because they provide daily outputs and are found to have a realistic performance over North America (@mahony22)

The benefits of using CMIP6 data are that each model in the ensemble a) has been rigorously evaluated, and b) uses the same standard socioeconomic trajectories—known as Shared Socioeconomic Pathways (SSPs)—to describe future greenhouse gas emissions. "The SSPs are based on five narratives describing alternative socio-economic developments, including sustainable development, regional rivalry, inequality, fossil-fueled development, and middle-of-the-road development" (@ssp)

::: {.callout-note icon=false}
### Shared Socioeconomic Pathways {#ssp-callout}
There are five different SSP categories that climate projections are grouped by (following text taken from @ssp): 
- SSP1	Sustainability: Taking the Green Road (Low challenges to mitigation and adaptation)
The world shifts gradually, but pervasively, toward a more sustainable path, emphasizing more inclusive development that respects perceived environmental boundaries. Management of the global commons slowly improves, educational and health investments accelerate the demographic transition, and the emphasis on economic growth shifts toward a broader emphasis on human well-being. Driven by an increasing commitment to achieving development goals, inequality is reduced both across and within countries. Consumption is oriented toward low material growth and lower resource and energy intensity.

- SSP2 Middle of the Road: (Medium challenges to mitigation and adaptation)
The world follows a path in which social, economic, and technological trends do not shift markedly from historical patterns. Development and income growth proceeds unevenly, with some countries making relatively good progress while others fall short of expectations. Global and national institutions work toward but make slow progress in achieving sustainable development goals. Environmental systems experience degradation, although there are some improvements and overall the intensity of resource and energy use declines. Global population growth is moderate and levels off in the second half of the century. Income inequality persists or improves only slowly and challenges to reducing vulnerability to societal and environmental changes remain.

- SSP3 Regional Rivalry: A Rocky Road (High challenges to mitigation and adaptation)
A resurgent nationalism, concerns about competitiveness and security, and regional conflicts push countries to increasingly focus on domestic or, at most, regional issues. Policies shift over time to become increasingly oriented toward national and regional security issues. Countries focus on achieving energy and food security goals within their own regions at the expense of broader-based development. Investments in education and technological development decline. Economic development is slow, consumption is material-intensive, and inequalities persist or worsen over time. Population growth is low in industrialized and high in developing countries. A low international priority for addressing environmental concerns leads to strong environmental degradation in some regions.

- SSP4 Inequality: A Road Divided (Low challenges to mitigation, high challenges to adaptation)
Highly unequal investments in human capital, combined with increasing disparities in economic opportunity and political power, lead to increasing inequalities and stratification both across and within countries. Over time, a gap widens between an internationally-connected society that contributes to knowledge- and capital-intensive sectors of the global economy, and a fragmented collection of lower-income, poorly educated societies that work in a labor intensive, low-tech economy. Social cohesion degrades and conflict and unrest become increasingly common. Technology development is high in the high-tech economy and sectors. The globally connected energy sector diversifies, with investments in both carbon-intensive fuels like coal and unconventional oil, but also low-carbon energy sources. Environmental policies focus on local issues around middle and high income areas.

- SSP5	Fossil-fueled Development: Taking the Highway (High challenges to mitigation, low challenges to adaptation)
This world places increasing faith in competitive markets, innovation and participatory societies to produce rapid technological progress and development of human capital as the path to sustainable development. Global markets are increasingly integrated. There are also strong investments in health, education, and institutions to enhance human and social capital. At the same time, the push for economic and social development is coupled with the exploitation of abundant fossil fuel resources and the adoption of resource and energy intensive lifestyles around the world. All these factors lead to rapid growth of the global economy, while global population peaks and declines in the 21st century. Local environmental problems like air pollution are successfully managed. There is faith in the ability to effectively manage social and ecological systems, including by geo-engineering if necessary.
:::

For the Montana Climate Assessment, we explore the SSP1, SSP2, SSP3, and SSP5 scenarios. 

Due to their complexity and global extent, GCMs can be computationally intensive. Thus, scientists often make climate projections at coarse spatial resolution where each projected data point is an average value of a grid cell that measures hundreds of miles (kilometers) across. 

For areas where the terrain and land cover are relatively homogenous (e.g., an expanse of the Great Plains), such coarse grid cells may be adequate to capture important climate processes. But in areas with complex landscapes like Montana, data points so widely spaced are inadequate to reflect variability in terrain and vegetation and their influence on climate. A 100 mile (161 km) grid, for example, might not capture the climate effects of a small mountain range rising out of the eastern Montana plains or the climate differences between mountain summits and valleys in western Montana where temperature and precipitation vary greatly.

To capture such important terrain characteristics, scientist take the coarse-resolution output from a GCM and statistically attribute the results from those models to smaller regions at higher resolution (e.g., grid points at 1 mile rather than 100 mile apart). This process, called downscaling, more accurately represents climate across smaller, more complex landscapes, including Montana.

For this climate assessment, we used a statistical downscaling method called the Multivariate Adaptive Constructive Analogs. By using a downscaled dataset—rather than the original output from the ensemble of GCMs—we gained the ability to evaluate temperature and precipitation at relatively high resolution statewide before conveying the results at the climate division scale. Additionally, we were able to aggregate data points within each of Montana’s seven climate divisions (Figure 2-3), and look at Montana’s climate future in different geographic areas. Aggregating to the climate-division level minimizes the potential for false precision by reporting results at spatial scales that better represent underlying climate processes. 

The 8 downscaled GCMs in CMIP6 were evaluated at two future time periods: 1) mid century (2040–2069) and 2) end-of-century (2070–2099). Thirty-year averages of these future projections were then compared to a historical (1981–2010) 30-year average, which results in a projected difference, or change, from historical conditions. We make those projections using the stabilization the four SSPs outlined above. These future projections were then compared to the historical trends in Montana to reveal the major climate-associated changes that Montana is likely to experience in the future.

## Temperature Projections
```{r prep_temp, echo=FALSE, warning=FALSE, message=FALSE}
library(magrittr)

tmp_box_data <- normals::make_boxplot_data(
    stringr::str_subset(cmip_files, "tas.rds"), 
    shp = shp, 
    attr_id = params$attr_id
)

tmmx_map_data <- normals::make_map_data(
    stringr::str_subset(cmip_files, "tasmax.rds"), 
    shp = shp, 
    proj = normals::mt_state_plane
) 

tmmn_map_data <- normals::make_map_data(
    stringr::str_subset(cmip_files, "tasmin.rds"), 
    shp = shp, 
    proj = normals::mt_state_plane
) 

tavg_heat_data <- normals::make_heatmap_data(
    stringr::str_subset(cmip_files, "tas.rds"), 
    shp = shp, 
    attr_id = params$attr_id
)

abv90_map_data <- normals::make_map_data(
    stringr::str_subset(cmip_files, "above90.rds"), 
    shp = shp, 
    proj = normals::mt_state_plane,
    fun = "sum"
)

abv90_box_data <- normals::make_boxplot_data(
    stringr::str_subset(cmip_files, "above90.rds"), 
    shp = shp, 
    attr_id = params$attr_id,
    fun = "sum"
)

ff_map_data <- normals::make_map_data(
    stringr::str_subset(cmip_files, "freeze-free.rds"), 
    shp = shp, 
    proj = normals::mt_state_plane,
    fun = "sum"
)

ff_box_data <- normals::make_boxplot_data(
    stringr::str_subset(cmip_files, "freeze-free.rds"), 
    shp = shp, 
    attr_id = params$attr_id,
    fun = "sum"
) 
```

Below we provide projections for various aspects of Montana’s future temperature based on our modeling analysis.

We discuss a subset of our modeling results here, including a) temperature projections reported by the median values of the 8 GCM ensemble and b) figures that include maps and graphs that represent the median value and distribution of values observed for temperature across the 8 GCMs. 

### Average Annual Temperatures
@fig-tavg-box shows projected changes in average annual temperatures across all `r params$attr_plural` for both the mid- and end-of-century. Below, projected changes in average annual temperature across the domain and associated model agreements are given:

```{r, echo=FALSE,message=FALSE,results="asis"}
cat(make_bullet_text("mid", "tas", "°F", "average annual temperature"))
```

```{r, echo=FALSE,message=FALSE,results="asis"}
cat(make_bullet_text("end", "tas", "°F", "average annual temperature"))
```

```{r temp_box, echo=FALSE, message=FALSE}
#| label: fig-tavg-box
#| fig-width: 10
#| fig-height: 7
#| fig-cap: "Graphs showing the minimum, maximum, and median temperature increases (°F) projected for each climate division in both stabilization (RCP4.5) and business-as-usual (RCP8.5) emission scenarios. The top row shows mid-century (2040-2069) projections and the bottom row shows end-of-century (2070-2099) projections. The outline of each box is determined by model agreement on the sign of the change (positive or negative). A black outline means there is >=80% model agreement and a red outline means that there is <80% model agreement. In this case, all models indicated the direction of the temperature trend at an agreement of greater than 80%."

normals::make_boxplot_plot(
  tmp_box_data,
  "Temperature Change (°F)", 
  "Change in Mean Annual Temperature"
)
```

### Average Daily Minimum Temperatures
@fig-tmmn-map shows spatially distributed changes in minimum annual temperature across all `r params$attr_plural` for both the mid- and end-of-century. Below, projected changes in both minimum temperature and associated model agreements are given for the entire domain:

```{r, echo=FALSE,message=FALSE,results="asis"}
cat(make_bullet_text("mid", "tasmin", "°F", "average annual minimum temperature"))
```

```{r, echo=FALSE,message=FALSE,results="asis"}
cat(make_bullet_text("end", "tasmin", "°F", "average annual minimum temperature"))
```

```{r tmmn_map, echo=FALSE, warning=FALSE, message=FALSE}
#| label: fig-tmmn-map
#| fig.width: 10
#| fig.height: 15
#| fig-cap: "The projected increase in annual average daily minimum temperature (°F) for each climate division in Montana for the periods 2049-2069 and 2070-2099 for (A) stabilization (RCP4.5) and (B) business-as-usual (RCP8.5) emission scenarios."

normals::apply_map_by_scenario(
  dat = tmmn_map_data,
  shp = shp,
  hot = TRUE,
  title = "Change in Annual Minimum Temperature (°F)"
)
```

### Average Daily Maximum Temperatures
@fig-tmmx-map shows spatially distributed changes in maximum annual temperature across all `r params$attr_plural` for both the mid- and end-of-century. Below, projected changes in both maxiimum temperature and associated model agreements are given for the entire domain:

```{r, echo=FALSE,message=FALSE,results="asis"}
cat(make_bullet_text("mid", "tasmax", "°F", "average annual maximum temperature"))
```

```{r, echo=FALSE,message=FALSE,results="asis"}
cat(make_bullet_text("end", "tasmax", "°F", "average annual maximum temperature"))
```

```{r tmmx_map, echo=FALSE, warning=FALSE, message=FALSE}
#| label: fig-tmmx-map
#| fig.width: 10
#| fig.height: 15
#| fig-cap: "The projected increase in annual average daily maximum temperature (°F) for each climate division in Montana for the periods 2049-2069 and 2070-2099 for (A) stabilization (RCP4.5) and (B) business-as-usual (RCP8.5) emission scenarios."

normals::apply_map_by_scenario(
  dat = tmmx_map_data,
  shp = shp,
  hot = TRUE,
  title = "Change in Annual Maximum Temperature (°F)"
)
```

### Average Monthly Temperatures
@fig-tavg-heatmap shows projected changes in average monthly temperatures across all `r params$attr_plural` for both the mid- and end-of-century. Below, projected changes in average monthly temperature across the domain and associated model agreements are given:

- **fill in** 

@fig-tavg-heatmap shows projected monthly changes in temperature for both mid- and end-of-century. 
Average monthly temperatures are projected to increase across all climate divisions by mid century (2040-2069) and for both stabilization and business-as-usual emission scenarios (Figure 2-11). Average monthly temperatures in summer and winter generally show larger projected increases than those in spring and fall. In the business-as-usual emission scenario, August has the largest projected change across all climate divisions. 

```{r tavg_heatmap, echo=FALSE, warning=FALSE, message=FALSE}
#| label: fig-tavg-heatmap
#| fig.width: 10
#| fig.height: 15
#| fig-cap: " The projected monthly increase in average temperature (°F) for each climate division in Montana in the mid-century projections (2040-2069) for the (A) stabilization (RCP4.5) and (B) business-as-usual (RCP8.5) emission scenarios."
normals::apply_heatmap_by_scenario(
  dat = tavg_heat_data,
  hot = TRUE,
  title = "Monthly Change in Average Temperature (°F)"
)
```

### Number of Days Above 90°F (32°C)
@fig-90-map and @fig-90-box show projected changes in number of days above 90°F for both the mid- and end-of-century. Below, projected changes in days above 90°F across the domain and associated model agreements are given:

```{r, echo=FALSE,message=FALSE,results="asis"}
cat(make_bullet_text("mid", "above90", "days", "number of days above 90°F"))
```

```{r, echo=FALSE,message=FALSE,results="asis"}
cat(make_bullet_text("end", "above90", "days", "number of days above 90°F"))
```


```{r abv90_map, echo=FALSE, warning=FALSE, message=FALSE}
#| label: fig-90-map
#| fig.width: 10
#| fig.height: 15
#| fig-cap: "The projected increases in number of days above 90°F (32°C) for each climate division in Montana over two periods 2040-2069 and 2070-2099 for (A) stabilization (RCP4.5) and (B) business-as-usual (RCP8.5) emission scenarios."

normals::apply_map_by_scenario(
  dat = abv90_map_data,
  hot = TRUE,
  shp = shp,
  title = "Change in Annual Number of Days Above 90°F"
)
```

```{r abv90_box, echo=FALSE, warning=FALSE, message=FALSE}
#| label: fig-90-box
#| fig.width: 10
#| fig.height: 7
#| fig-cap: "Graphs showing the increase in the number of days per year above 90°F (32°C) projected for each climate division in both stabilization (RCP4.5) and business-as-usual (RCP8.5) emission scenarios. The top row shows mid-century projections (2040-2069) and the bottom row shows end-of-century projections (2070-2099). The outline of each box is determined by model agreement on the sign of the change (positive or negative). A black outline means there is ≥80% model agreement and a red outline means that there is <80% model agreement. In this case, all models indicated the direction of the trend for days above 90°F (32°C) at an agreement of greater than 80%."

normals::make_boxplot_plot(
  dat = abv90_box_data,
  ylab = "Number of Days",
  title = "Change in Number of Days Above 90°F"
)
```

### Number of Days Where Minimum Temperatures are Above 32°F (0°C)
@fig-ff-map and @fig-ff-box show projected changes in number of days freeze-free days for both the mid- and end-of-century. Below, projected changes in the number of freeze-free days across the domain and associated model agreements are given:

```{r, echo=FALSE,message=FALSE,results="asis"}
cat(make_bullet_text("mid", "freeze-free", "days", "number of freeze-free days"))
```

```{r, echo=FALSE,message=FALSE,results="asis"}
cat(make_bullet_text("end", "freeze-free", "days", "number of freeze-free days"))
```


```{r ff_map, echo=FALSE, warning=FALSE, message=FALSE}
#| label: fig-ff-map
#| fig.width: 10
#| fig.height: 15
#| fig-cap: "The projected change in the number of frost-free days for each climate division in Montana over two periods 2040-2069 and 2070-2099 for (A) stabilization (RCP4.5) and (B) business-as-usual (RCP8.5) emission scenarios."

normals::apply_map_by_scenario(
  dat = ff_map_data,
  hot = TRUE,
  shp = shp,
  title = "Change in Number of Freeze Free Days"
)
```

```{r ff_box, echo=FALSE, warning=FALSE, message=FALSE}
#| label: fig-ff-box
#| fig.width: 10
#| fig.height: 7
#| fig-cap: "Graphs showing the increases in frost-free days/yr projected for each climate division in both stabilization (RCP4.5) and business-as-usual (RCP8.5) emission scenarios. The top row shows mid-century projections (2040-2069) and the bottom row shows end-of-century projections (2070-2099). The outline of each box is determined by model agreement on the sign of the change (positive or negative). A black outline means there is >=80% model agreement and a red outline means that there is <80% model agreement. In this case, all models indicated the direction of the trend of frost-free days at an agreement of greater than 80%."

normals::make_boxplot_plot(
  dat = ff_box_data,
  ylab = "Number of Days",
  title = "Change in Number of Freeze Free"
)
```

### Summary 
**In general, there is high model agreement and low uncertainty that temperatures and associated temperature metrics will increase both by mid century and end-of-century. For both periods, annual and seasonal temperature averages, the number of days/yr with extreme heat, and the overall length of the growing season are projected to increase. Differences exist in projections for the stabilization and business-as-usual emission scenarios, with the former consistently showing lower magnitudes of change than the latter. Many of the trends and spatial patterns seen in the mid-century projections are extended and exacerbated in the end-of-century projections. The range of model outputs also increases for end-of-century projections, suggesting that the magnitude of change becomes more uncertain in the models further out in time.**

**Regardless of uncertainties, the GCMs show full agreement regarding the direction of change: temperatures will be increasing.**

## Precipitation Projections
Below we provide projections of `r params$location`’s future precipitation based on our modeling efforts. Those projections cover all SSP scenarios and two periods: mid century (2040-2069) and end-of-century (2070-2099).

We discuss a subset of our precipitation modeling results here, including a) precipitation projections reported by the mean values of the 8 GCM ensemble and b) figures that include maps and graphs that represent the mean and distribution of values observed for precipitation across the 8 GCMs. Special consideration is required for interpretations of precipitation changes in Montana’s complex terrain. Precipitation increases drastically with elevation such as that found in northwest Montana. Here, mean values do not characterize the potential for spatial variability that exists within these regions.

### Average Annual Precipitation
@fig-pr-map and @fig-pr-box show projected changes in annual total precipitation for both the mid- and end-of-century. Below, projected changes in total annual precipitation across the domain and associated model agreements are given:

```{r, echo=FALSE,message=FALSE,results="asis"}
cat(make_bullet_text("mid", "pr", "inches", "total annual precipitation"))
```

```{r, echo=FALSE,message=FALSE,results="asis"}
cat(make_bullet_text("end", "pr", "inches", "total annual precipitation"))
```


```{r pr_map, echo=FALSE, warning=FALSE, message=FALSE}
#| label: fig-pr-map
#| fig.width: 10
#| fig.height: 15
#| fig-cap: "The projected change in annual precipitation (inches) for each climate division in Montana over two periods 2040-2069 and 2070-2099 for (A) stabilization (RCP4.5) and (B) business-as-usual (RCP8.5) emission scenarios."

pr_map_data <- normals::make_map_data(
    f = stringr::str_subset(cmip_files, "pr.rds"), 
    shp = shp, 
    proj = normals::mt_state_plane,
    fun = "sum"
) 

normals::apply_map_by_scenario(
  dat = pr_map_data,
  hot = TRUE,
  shp = shp,
  title = "Change in Total Annual Precipitation (in.)"
)
```

```{r pr_box, echo=FALSE, warning=FALSE, message=FALSE}
#| label: fig-pr-box
#| fig.width: 10
#| fig.height: 7
#| fig-cap: "Graphs showing annual precipitation change (in inches) projected for each climate division in both stabilization (RCP4.5) and business-as-usual (RCP8.5) emission scenarios. The top row shows mid-century projections (2040-2069) and the bottom row shows end-of-century projections (2070-2099). The outline of each box is determined by model agreement on the sign of the change (positive or negative). A black outline means there is >=80% model agreement and a red outline means that there is <80% model agreement. In this case, all models indicated the direction of the annual precipitation trend at an agreement of greater than 80%."

pr_box_data <- normals::make_boxplot_data(
    stringr::str_subset(cmip_files, "pr.rds"), 
    shp = shp, 
    attr_id = params$attr_id,
    fun = "sum"
) 

normals::make_boxplot_plot(
  dat = pr_box_data,
  ylab = "Precipitation (in.)",
  title = "Change in Monthly Precipitation (in.)"
)
```
### Interannual Variability
Interannual variability (i.e., the amount precipitation changes from year to year) is projected to `r subset_agreements("iv", "mid_century", fun=most_common, col="txt")` across the domain by mid century and `r subset_agreements("iv", "mid_century", fun=most_common, col="txt")` by the end-of-century for the majority of SSP scenarios (@fig-iv-box). These changes could be attributed to wet years getting wetter, dry years getting drier, or some combination of both.

```{r iv_box, echo=FALSE, warning=FALSE, message=FALSE}
#| label: fig-iv-box
#| fig.width: 10
#| fig.height: 7
#| fig-cap: "Graphs showing the interannual variability of precipitation projected for each climate division in both stabilization (RCP4.5) and business-as-usual (RCP8.5) emission scenarios. The top row shows mid-century projections (2040-2069) and the bottom row shows for end-of-century projections (2070-2099). The outline of each box is determined by model agreement on the sign of the change (positive or negative). A black outline means there is ≥80% model agreement and a red outline means that there is <80% model agreement."

iv_box_data <- normals::make_boxplot_data(
    stringr::str_subset(cmip_files, "iv.rds"), 
    shp = shp, 
    attr_id = params$attr_id,
    fun = "mean"
) 

normals::make_boxplot_plot(
  dat = iv_box_data,
  ylab = "Inches",
  title = "Change in Interannual Precipitation Variability (in.)"
)
```
### Monthly and Seasonal Change in Average Precipitation
**FILL IN While annual increases in precipitation are projected across the state with moderately high model agreement, the monthly and seasonal projections vary. In mid-century projections, winter, spring, and fall increase in monthly precipitation for both emission scenarios, with spring experiencing the largest increases (e.g., 0.4 inch/month [1.0 cm/month] for the business-as-usual emission scenario; Figure 2-23). Summers, however, are projected to decrease by about 0.1 inch/month (0.3 cm/month) in both emission scenarios (model agreement, however, is fairly low for these projections). For end-of-century projections, the same trends are seen for increasing precipitation in winter, spring, and fall and decreasing precipitation in summer. The magnitude of change is similar to that of mid-century projections.**

```{r pr_heatmap, echo=FALSE, warning=FALSE, message=FALSE}
#| label: fig-pr-heatmap
#| fig.width: 10
#| fig.height: 15
#| fig-cap: "Projected monthly change in average precipitation (inches) for each climate division in Montana in the mid-century projections (2040-2069) for (A) stabilization (RCP4.5) and (B) business-as-usual (RCP8.5) emission scenarios."

pr_heat_data <- normals::make_heatmap_data(
    stringr::str_subset(cmip_files, "pr.rds"), 
    shp = shp, 
    attr_id = params$attr_id
) 

normals::apply_heatmap_by_scenario(
  dat = pr_heat_data,
  hot = FALSE,
  title = "Monthly Change in Total Precipitation (in.)"
)
```

- **Mid-century projection specifics.—Although annual precipitation increases across the state with moderately high model agreement, the monthly and seasonal projections vary somewhat. Winter, spring, and fall increase in monthly precipitation for both emission scenarios, with the largest increases in spring (Figure 2-20). For the stabilization emission scenario, spring months increase by about 0.2 inch/month (0.5 cm/month) (minimum: -0.1 inch/month [-0.3 cm/month], maximum: 0.8 inch/month [2.0 cm/month], percent agreement: 85%). In the business-as-usual emission scenario, spring months increase by 0.4 inch/month (1.0 cm/month) (minimum: 0.0 inch/month [0 cm/month], maximum: 1.0 inch/month [2.5 cm/month], percent agreement: 95%). Summer months, however, show decreasing precipitation for both scenarios, although model agreement is fairly low in the projections. For the both the stabilization and business-as-usual emission scenarios, summer precipitation decreases by -0.1 inch/month (-0.3 cm/month) (minimum: -0.4 inch/month [-1.0 cm/month], maximum: 0.5 inch/month [1.3 cm/month], percent agreement: 65%). **
- **Mid-century projection specifics.—Although annual precipitation increases across the state with moderately high model agreement, the monthly and seasonal projections vary somewhat. Winter, spring, and fall increase in monthly precipitation for both emission scenarios, with the largest increases in spring (Figure 2-20). For the stabilization emission scenario, spring months increase by about 0.2 inch/month (0.5 cm/month) (minimum: -0.1 inch/month [-0.3 cm/month], maximum: 0.8 inch/month [2.0 cm/month], percent agreement: 85%). In the business-as-usual emission scenario, spring months increase by 0.4 inch/month (1.0 cm/month) (minimum: 0.0 inch/month [0 cm/month], maximum: 1.0 inch/month [2.5 cm/month], percent agreement: 95%). Summer months, however, show decreasing precipitation for both scenarios, although model agreement is fairly low in the projections. For the both the stabilization and business-as-usual emission scenarios, summer precipitation decreases by -0.1 inch/month (-0.3 cm/month) (minimum: -0.4 inch/month [-1.0 cm/month], maximum: 0.5 inch/month [1.3 cm/month], percent agreement: 65%). **

## Projected Changes in Consecutive Dry Days
To assess changes in the frequency of dry events, we determined the annual number of dry days (defined as days when precipitation is less than 0.01 inch [0.03 cm]), then calculated the maximum number of consecutive dry days/yr averaged over the 30-year periods of interest. Across the domain, we found a `r subset_agreements("con-dry", "mid_century", fun = most_common, col = "txt")` in consecutive dry days for the mid-century across the all SSPs (`r subset_agreements("con-dry", "mid_century", fun = mean, col = "agree")`% agreement) and a `r subset_agreements("con-dry", "end_century", fun = most_common, col = "txt")` in consecutive dry days for the end-of-century across the all SSPs (`r subset_agreements("con-dry", "end_century", fun = mean, col = "agree")`% agreement). Figures @fig-dd-map and @fig-dd-box show changes in projected number of dry days across `r params$attr_plural` and domain-wide projections for mid- and end-of-century are given below: 

```{r, echo=FALSE,message=FALSE,results="asis"}
cat(make_bullet_text("mid", "con-dry", "days", "annual number of consecutive dry days"))
```

```{r, echo=FALSE,message=FALSE,results="asis"}
cat(make_bullet_text("end", "con-dry", "days", "annual number of consecutive dry days"))
```


```{r dd_map, echo=FALSE, warning=FALSE, message=FALSE}
#| label: fig-dd-map
#| fig.width: 10
#| fig.height: 15
#| fig-cap: "The projected change in the number of consecutive dry days (<0.1 inch [0.3 cm] of precipitation) for each climate division in Montana over two periods 2040-2069 and 2070-2099 for (A) stabilization (RCP4.5) and (B) business-as-usual (RCP8.5) emission scenarios."

dd_map_data <- normals::make_map_data(
    f = stringr::str_subset(cmip_files, "con-dry.rds"), 
    shp = shp, 
    proj = normals::mt_state_plane,
    fun = "sum"
) 

normals::apply_map_by_scenario(
  dat = dd_map_data,
  hot = FALSE,
  shp = shp,
  title = 'Change in Number of Consecutive Dry Days (<0.1in")'
)
```

```{r dd_box, echo=FALSE, warning=FALSE, message=FALSE}
#| label: fig-dd-box
#| fig.width: 10
#| fig.height: 7
#| fig-cap: "Graphs showing the number of consecutive dry days in a year projected for each climate division in both stabilization (RCP4.5) and business-as-usual (RCP8.5) emission scenarios. The top row shows mid-century projections (2040-2069) and the bottom row shows end-of-century projections (2070-2099). The outline of each box is determined by model agreement on the sign of the change (positive or negative). A black outline means there is ≥80% model agreement and a red outline means that there is <80% model agreement. In the case of consecutive dry days, there was less than 80% agreement across the models for all climate divisions."

dd_box_data <- normals::make_boxplot_data(
    stringr::str_subset(cmip_files, "con-dry.rds"), 
    shp = shp, 
    attr_id = params$attr_id,
    fun = "sum"
) 

normals::make_boxplot_plot(
  dat = dd_box_data,
  ylab = "Nunber of Days",
  title = 'Change in Number of Consecutive Dry Days (<0.1in")'
)
```

### Projected Change in Wet Days
To evaluate changes in wet events, we calculated the number of days/yr where precipitation is greater than 1.0 inch (2.5 cm) and average those values over the period of interest. Figures @fig-wd-box shows changes in projected number of wet days across `r params$attr_plural` and domain-wide projections for mid- and end-of-century are given below: 

```{r, echo=FALSE,message=FALSE,results="asis"}
cat(make_bullet_text("mid", "con-wet", "days", "annual number of consecutive wet days"))
```

```{r, echo=FALSE,message=FALSE,results="asis"}
cat(make_bullet_text("end", "con-wet", "days", "annual number of consecutive wet days"))
```

```{r wd_box, echo=FALSE, warning=FALSE, message=FALSE}
#| label: fig-wd-box
#| fig.width: 10
#| fig.height: 7
#| fig-cap: "Graphs showing the increase in the number of wet days/yr projected for each climate division in both stabilization (RCP4.5) and business-as-usual (RCP8.5) emission scenarios. The top row shows projections for mid century (2040-2069) and the bottom row shows projections for end-of-century (2070-2099). The outline of each box is determined by model agreement on the sign of the change (positive or negative). A black outline means there is ≥80% model agreement and a red outline indicates <80% model agreement. Model agreement for the trend of wet days each year was greater than 80%, except for the northeastern climate division."

wd_box_data <- normals::make_boxplot_data(
    stringr::str_subset(cmip_files, "con-wet.rds"), 
    shp = shp, 
    attr_id = params$attr_id,
    fun = "sum"
) 

normals::make_boxplot_plot(
  dat = wd_box_data,
  ylab = "Nunber of Days",
  title = 'Change in Number of Consecutive Wet Days (>1 in")'
)
```

### Summary
In mid-century and end-of-century projections, average annual precipitation and variability increase across the state, as does winter, spring, and fall precipitation. Summers, however, show slight decreases in precipitation. The projections suggest little change in the annual frequency of dry and wet events, although there is high uncertainty in the case of wet events. Similar analysis using different metrics for the larger region surrounding Montana indicates an even larger potential (30%) for more days of extreme precipitation (NCA 2014). Overall, the differences in precipitation resulting from the different emission scenarios (i.e., stabilization versus business-as-usual) are small when compared to the impact of the emission scenarios on the temperature projections. Uncertainty in the projections generally increases the further out in time (i.e., in the end-of-century projections), as well as for the higher business-as-usual emission scenario.

