---
params: 
  location: "Montana"
  asset_dir: "./assets"
  attr_id: "Division"
  attr_name: "Climate Division"
  attr_plural: "Climate Divisions"
  
---

```{r setup, echo=FALSE}
library(magrittr)

shp <- sf::read_sf(file.path(params$asset_dir, "agg_regions.geojson"))
whole <- shp %>% 
  dplyr::group_by(out = params$attr_id) %>% 
  dplyr::summarise() %>%
  dplyr::transmute(region = "region")


rasts <- list.files(params$asset_dir, pattern = "nclimgrid_historical_seasonal.tif", full.names = T) %>%
  tibble::tibble(f = .) %>% 
  dplyr::rowwise() %>% 
  dplyr::mutate(r = list(terra::rast(f) %>% terra::mask(shp))) %>%
  dplyr::mutate(variable = basename(f) %>% 
                    stringr::str_split("_") %>% 
                    unlist() %>% 
                    head(1))

historical_table <- 
  purrr::map(rasts$f, function(x) {
    x %>% 
      terra::rast() %>% 
      normals::spat_summary(shp, params$attr_id, "season", "mean", na.rm = T) %>%
      dplyr::mutate(f = x)
  }) %>% 
    dplyr::bind_rows() %>%
    dplyr::rowwise() %>% 
    dplyr::mutate(variable = basename(f) %>% 
                    stringr::str_split("_") %>% 
                    unlist() %>% 
                    head(1)
    ) %>% 
    dplyr::ungroup() %>%
    dplyr::select(Division, season, value, variable) %>% 
    dplyr::mutate(
      value = ifelse(variable == "pr", value / 25.4, (value) * 9/5 + 32)
    )

global_avg <- rasts %>% 
  dplyr::mutate(r = list(terra::global(r, fun="mean", na.rm = T) %>%
                           tibble::rownames_to_column() %>%
                           dplyr::rename(season=rowname))) %>%
  tidyr::unnest(r) %>% dplyr::select(value=mean, variable, season) %>%
  dplyr::mutate(
    value = ifelse(variable == "pr", value / 25.4, (value) * 9/5 + 32)
  )

annual_tavg <- rasts %>% 
  dplyr::filter(variable == "tavg") %$% 
  terra::mean(r[[1]]) %>% 
  {(.) * 9/5 + 32}

east_west_ppt <- rasts %>% 
  dplyr::filter(variable == "pr") %$%
  terra::app(r[[1]], fun = "sum") %>%
  {./25.4} %>%
  normals::spat_summary(
    shp = normals::divide, attr_id = "loc", name_to = "sum", fun = "mean"
  ) 

monthly_tmp <- terra::rast("./assets/tavg_nclimgrid_monthly.tif") %>% 
  {(. * 9/5) + 32} %>% terra::mask(whole)
monthly_tmmx <- terra::rast("./assets/tmmx_nclimgrid_monthly.tif") %>% 
  terra::mask(whole)
monthly_tmmn <- terra::rast("./assets/tmmn_nclimgrid_monthly.tif")  %>% 
  terra::mask(whole)
monthly_ppt <- terra::rast("./assets/pr_nclimgrid_monthly.tif") %>% 
  {. / 25.4} %>% terra::mask(whole)
annual_tmp <- monthly_tmp %>% 
  terra::tapp(fun = "mean", index = "years") %>% 
  terra::`time<-`(as.Date(glue::glue("{names(.)}0101"), format = "X%Y%m%d"))
```

# **Draft** Historical Climate {#sec-historical}
## Climate Conditions 1991 - 2020
To assess `r params$location`’s current climate, we analyzed climate variable data (see sidebar) from the NClimGrid climate dataset. In this section, we review average temperature and precipitation conditions from 1991-2020 as an indicator of current climate conditions.

::: {.callout-note icon=false}
### Climate Variables {#climate-callout}

In analyses of climate, scientists employ a suite of 50 essential climate variables to unify discussions (Global Climate Observing System undated). For this assessment, we primarily focus on just two: how climate change will affect Montana’s temperature and precipitation in the future. 

- Temperature is an objective measure of how hot or cold and object is with reference to some standard value. Temperature differences across the Earth result primarily from regional differences in absorbed solar radiation. Seasonal variations in temperature result from the tilt of the Earth’s axis as it rotates around the sun. 

- Precipitation is the quantity of water (solid or liquid) falling to the Earth’s surface at a specific place during a given period. Like temperature, precipitation varies seasonally and from place to place. Precipitation amounts can have a dramatic impact on local environmental conditions, such as abundance of wildlife or potential for crop production.
:::

### Temperature
@tbl-historical-temp shows the average seasonal temperature variation across `r params$location`’s `r nrow(shp)` `r params$attr_plural` from 1981-2010. Temperatures vary across `r params$location`, with average annual values ranging from `r terra::global(annual_tavg, fun="min", na.rm = T) %>% as.numeric() %>% round(1)`°F to `r terra::global(annual_tavg, fun="max", na.rm = T) %>% as.numeric() %>% round(1)`°F across the region. 

```{r echo=FALSE}
#| label: tbl-historical-temp
#| tbl-cap: "Average (minimum / average / maximum) temperatures (°F) for the seven Montana climate divisions from 1981-2010."
historical_table %>% 
  dplyr::mutate(value = round(value, 1)) %>%
  tidyr::pivot_wider(names_from = variable, values_from = value) %>% 
  dplyr::mutate(value = glue::glue("{tmmn} / {tavg} / {tmmx}") %>% 
                  as.character()) %>% 
  dplyr::select(Division, season, value) %>% 
  tidyr::pivot_wider(names_from = season, values_from = value) %>% 
  knitr::kable()
```

Seasonal temperatures across `r params$location` vary, with seasonal average Winter, Spring, Summer and Fall temperatures of `r dplyr::filter(global_avg, variable == "tavg", season == "Winter") %$% round(value, 1)`°F, `r dplyr::filter(global_avg, variable == "tavg", season == "Spring") %$% round(value, 1)`°F, `r dplyr::filter(global_avg, variable == "tavg", season == "Summer") %$% round(value, 1)`°F, and `r dplyr::filter(global_avg, variable == "tavg", season == "Fall") %$% round(value, 1)`°F, respectively. 

### Precipitation
@tbl-historical-pr shows the seasonal variation of precipitation across `r params$location`’s `r nrow(shp)` `r params$attr_plural` from 1981-2010. The average annual precipitation for `r params$location` is `r dplyr::filter(global_avg, variable == "pr") %$% sum(value) %>% round(1)`. 

```{r echo=FALSE}
#| label: tbl-historical-pr
#| tbl-cap: "Average precipitation in inches for the seven Montana climate divisions from 1981-2010."
historical_table %>% 
  dplyr::mutate(value = round(value, 1)) %>%
  dplyr::filter(variable == "pr") %>% 
  dplyr::select(-variable) %>% 
  tidyr::pivot_wider(names_from = season, values_from = value) %>% 
  knitr::kable() 
```
## Historical Trends 1951 - Present
We evaluated how temperature and precipitation have historically changed, dating back to 1950. This review of historical trends helps us provide context for future climate change scenarios explored in later sections of this chapter. In addition, evaluating these trends can help us better understand a) how `r params$location` has previously experienced and responded to changing climate, b) if projections of future change reveal a different climate than we have previously experienced, and c) the potential impacts of that projected change. 

The presentation of trends that follows is confined to the period from 1950-2020 using data from NOAA's NClimGrid Dataset (@nclimgrid). This is widely acknowledged as the benchmark period in climate analysis (@liebmann), a period when our network of meteorological sensors becomes more accurate and sufficiently dense.

### Temperature
@tbl-temp-trends shows the decadal rate of change from 1950-2020 for average annual temperatures across `r params$location`'s `r nrow(shp)` `r params$attr_plural`. We provide that rate of change both annually and by season for `r params$location`. We also present the average annual and average seasonal changes across the region. To account partially for autocorrelation we considered trends as significant with a conservative p value at p<0.05. 

```{r echo=FALSE}
#| label: tbl-temp-trends
#| tbl-cap: "Decadal rate of change for annual average temperatures in °F (°C) for each region in the study area from 1950-2020. A value of 0 indicates no statistically significant change between decadal averages."
by_div <- dplyr::full_join(
  normals::make_seasonal_trend_table(monthly_tmp, shp, "Division", "mean", 1951, 2020),
  normals::make_seasonal_trend_table(monthly_tmp, shp, "Division", "mean", 1951, 2020, TRUE),
  by="Division"
)

by_state <- dplyr::full_join(
  normals::make_seasonal_trend_table(monthly_tmp, whole, "region", "mean", 1951, 2020),
  normals::make_seasonal_trend_table(monthly_tmp, whole, "region", "mean", 1951, 2020, TRUE),
  by="region"
) %>%
  dplyr::mutate(Division="Statewide")

state_tmmx_trend <- 
  normals::make_seasonal_trend_table(monthly_tmmx, whole, "region", "mean", 1951, 2020, TRUE) %$%
  round(Annual, 2) 

state_tmmn_trend <- 
  normals::make_seasonal_trend_table(monthly_tmmn, whole, "region", "mean", 1951, 2020, TRUE) %$%
  round(Annual, 2)

state_tavg_trend <- 
  normals::make_seasonal_trend_table(monthly_tmp, whole, "region", "mean", 1951, 2020, TRUE) %$%
  round(Annual, 2)

all_tmp_trends <- dplyr::bind_rows(by_div, by_state) %>% 
  dplyr::select(Division, Annual, Winter, Spring, Summer, Fall) %>% 
  dplyr::mutate(dplyr::across(where(is.numeric), ~ round(.x, 2))) 

get_trend_direction <- function(x) {
  if (x == 0) {
    return("not changed")
  } else if (x > 0) {
    return("increased")
  } else {
    return("decreased")
  }
}

knitr::kable(all_tmp_trends)
```

The rate of temperature change across `r params$location` was `r dplyr::filter(all_tmp_trends, Division == "Statewide")  %$% round(Annual, 2)`°F/decade (@tbl-temp-trends). Across `r params$attr_plural` average annual minimum and maximum temperature changes ranged from `r state_tmmx_trend`-`r state_tmmn_trend`°F/decade, respectively. Between 1950 and 2020, `r params$location`’s average annual temperature has `r get_trend_direction(state_tavg_trend)` by `r dplyr::filter(all_tmp_trends, Division == "Statewide") %$% round(Annual * 6.9, 2)`°F; annual maximum and minimum temperatures have `r get_trend_direction(state_tmmx_trend)` by approximately `r round(state_tmmx_trend * 6.9, 2)`°F and `r round(state_tmmn_trend * 6.9, 2)`°F, respectively.

```{r echo=FALSE, warning=FALSE, message=FALSE, eval=TRUE}
#| label: fig-temp-trends
#| fig-cap: !expr glue::glue("Trends in annual average temperature across each {params$attr_name} in {params$location}")
#| fig-height: 7
#| fig-width: 10

normals::make_trend_fig(annual_tmp, shp, "Division", "mean", 1951, 2020, "Temperature °F")
```

### Precipitation
@tbl-ppt-trends shows the decadal rate of change from 1950-2020 for average total precipitation across `r params$location`'s `r nrow(shp)` `r params$attr_plural`. We provide that rate of change both annually and by season for `r params$location`. We also present the average annual and average seasonal changes across the region. To account partially for autocorrelation we considered trends as significant with a conservative p-value at p<0.05. 

```{r echo=FALSE}
#| label: tbl-ppt-trends
#| tbl-cap: "Decadal rate of change in average precipitation in inches/decade for the seven Montana climate divisions from 1951-2020. A value of 0 indicates no significant change."

by_div <- dplyr::full_join(
  normals::make_seasonal_trend_table(monthly_ppt, shp, "Division", "sum", 1951, 2020),
  normals::make_seasonal_trend_table(monthly_ppt, shp, "Division", "sum", 1951, 2020, TRUE),
  by="Division"
)

by_state <- dplyr::full_join(
  normals::make_seasonal_trend_table(monthly_ppt, whole, "region", "sum", 1951, 2020),
  normals::make_seasonal_trend_table(monthly_ppt, whole, "region", "sum", 1951, 2020, TRUE),
  by="region"
) %>%
  dplyr::mutate(Division="Statewide")

state_ppt_trend <- 
  normals::make_seasonal_trend_table(monthly_ppt, whole, "region", "sum", 1951, 2020, TRUE) %$%
  round(Annual, 2)

ppt_trends <- dplyr::bind_rows(by_div, by_state) %>% 
  dplyr::select(Division, Annual, Winter, Spring, Summer, Fall) %>% 
  dplyr::mutate(dplyr::across(where(is.numeric), ~ round(.x, 2))) 

knitr::kable(ppt_trends)
```

The rate of precipitation change across `r params$location` was `r dplyr::filter(ppt_trends, Division == "Statewide") %$% round(Annual, 2)`in./decade (@tbl-ppt-trends). Between 1950 and 2020, `r params$location`’s average annual precipitation has `r get_trend_direction(state_ppt_trend)` by `r dplyr::filter(ppt_trends, Division == "Statewide") %$% round(Annual * 6.9, 2)` inches.

## Extreme aspects of Montana’s climate 

```{r echo=FALSE, cache=TRUE, warning=FALSE, message=FALSE}
c_to_f <- function(x) (x * 9/5) + 32 
extremes <- readRDS("./assets/climdex.rds") %>% 
  dplyr::mutate(value = ifelse(stringr::str_detect(name, "monthly"), c_to_f(value), value)) %>% 
  dplyr::mutate(value = ifelse(stringr::str_detect(name, "monthly") && period == "diff", value - 32, value)) %>% 
  dplyr::mutate(value = round(value, 1))

cool_diff = extremes %>% dplyr::filter(name == "cool_days", period == "diff") %$% value 
icing_diff = extremes %>% dplyr::filter(name == "icing_days", period == "diff") %$% value 
growing_diff = extremes %>% dplyr::filter(name == "growing_season", period == "diff") %$% value 
warm_diff = extremes %>% dplyr::filter(name == "warm_days", period == "diff") %$% value 
mon_min_diff = extremes %>% dplyr::filter(name == "monthly_min", period == "diff") %$% value 
mon_max_diff = extremes %>% dplyr::filter(name == "monthly_max", period == "diff") %$% value 
```
Along with analyzing historical trends in temperature and precipitation, we performed an analysis of changes in extreme climate events since the middle of last century. Two examples of climate extremes include periods of intense warm or cool temperatures and significant wet or dry spells across seasons. Because these events affect every aspect of our society, decision makers and stakeholders are increasingly in need of historical evaluations of extreme events and how they are changing from seasons to centuries. Since 1951, our analysis shows the average winter temperature has increased by `r all_tmp_trends %>% dplyr::filter(Division == "Statewide") %$% Winter`°F/decade across the state, with an overall average winter temperature increase of `r all_tmp_trends %>% dplyr::filter(Division == "Statewide") %$% Winter * 6.9`°F. Average spring temperatures have increased by `r all_tmp_trends %>% dplyr::filter(Division == "Statewide") %$% Spring * 6.9`°F during the same period, and average summer temperatures have risen by `r all_tmp_trends %>% dplyr::filter(Division == "Statewide") %$% Summer * 6.9`°F. Montana’s fall average temperatures have increased by `r all_tmp_trends %>% dplyr::filter(Division == "Statewide") %$% Fall * 6.9`°F since 1951.

We performed our analysis of climate extremes using extreme climate indices outlined by the CLIMDEX project (CLIMDEX undated). The extreme climate indices were derived using NClimGrid data. Extreme precipitation events across the United States have increased in both intensity and frequency since 1901 (NCA 2014), including across both the High Plains and the northwestern US (many states combined), where studies have shown an increase in the number of days with extreme precipitation (NCA 2014). Here, we report change in temperature extremes for `r params$location` and, for perspective, the climate normals for these extremes for the periods 1951-1980 and from 1991-2020 (@tbl-xtremes).

```{r xt-table, echo=FALSE}
#| label: tbl-xtremes
#| tbl-cap: !expr glue::glue("Changes in {params$location}’s climate extremes. For historical perspective, we also report the climate normal for these extremes from the periods 1951-1980 and from 1991-2020.")

dplyr::select(extremes, name, period, value) %>% 
  dplyr::ungroup() %>% 
  dplyr::mutate(
    name = dplyr::recode(
      name,
      cool_days = "Frost Days",
      icing_days = "Icing Days",
      growing_season = "Growing Season Length (Days)",
      warm_days = "Summer Days",
      monthly_max = "Max Monthly Temperature (°F)",
      monthly_min = "Min Monthly Temperature (°F)"
    ),
    period = dplyr::recode(
      period,
      start = "1951-1980",
      end = "1991-2020",
      diff = "Change (1951-2020)"
    ) 
  ) %>% 
  tidyr::pivot_wider(names_from = period, values_from = value) %>% 
  dplyr::rename(Variable=name) %>% 
  knitr::kable()
```

The annual number of frost days and the number of icing days are `r extremes %>% dplyr::filter(name %in% c("icing_days", "cool_days"), period == "diff") %$% ifelse(mean(value) <= 0, "decreasing", "increasing")` across `r params$location` We use the CLIMDEX definition of frost days (icing days) as the number of days in a year when minimum (maximum) temperature is below 32°F. Coincident with warming temperatures, the number of frost days each year during the period from 1951-2010 has `r ifelse(cool_diff <= 0,"decreased", "increased")` by `r abs(cool_diff)` days. Along with this trend, the number of days in which the maximum temperatures are below 32°F (0°C; i.e. icing days) has `r ifelse(icing_diff <= 0,"decreased", "increased")` by `r abs(icing_diff)` days during this time period. These trends have contributed to an overall `r ifelse(growing_diff <= 0,"decrease", "increase")` in the growing season length of `r abs(growing_diff)` days between 1951 and 2010. In addition, the number of summer days, where maximum temperature exceeds 77°F (25°C) based on historical conditions, has `r ifelse(warm_diff <= 0,"decreased", "increased")` by `r abs(warm_diff)` days over this period. At a sub-annual level, monthly maximum and minimum temperatures have also changed. These are defined as the monthly maximum (minimum) value of daily maximum (minimum) temperatures. Monthly minimum values of daily minimum temperatures have `r ifelse(mon_min_diff <= 0,"decreased", "increased")` by `r abs(mon_min_diff)`°F from the period 1951-2010. Over the same time period, monthly minimum values of daily maximum temperatures have `r ifelse(mon_max_diff <= 0,"decreased", "increased")` by `r abs(mon_max_diff)`°F

::: {.callout-note icon=false}
### Drought {#drought-callout}

Drought is a recurrent climate event that may vary in intensity and persistence by region. Drought can have broad and potentially devastating environmental and economic impacts (Wilhite 2000); thus, it is a topic of ongoing, statewide concern.

Through time, Montana’s people, agriculture, and industry, like its ecosystems, have evolved with drought. Today, many entities across the state address drought, including private and non-profit organizations, state and federal agencies, and landowners, as well as unique watershed partnerships.

Drought is a complex phenomenon driven by both climate, but also affected by human-related factors (e.g., land use, water use). Although the definition of drought varies in different operational contexts, most definitions include several interrelated components, including:

- meteorological drought, defined as a deficit in precipitation and above average evapotranspiration that lead to increased aridity;

- hydrological drought, characterized by reduced water levels in streams, lakes, and aquifers following prolonged periods of meteorological drought;

- ecological drought, defined as a prolonged period over which an ecosystem’s demand for water exceeds the supply (the resulting water deficit, or shortage, creates multiple stresses within and across ecosystems); and

- agricultural drought, commonly understood as a deficit in soil moisture and water supply that lead to decreased productivity (in this assessment, we will treat this form of drought as an important component of ecological drought).

While the subsequent chapters dealing with water, agriculture, and forests treat the subject of drought differently, each describes drought within the context of one or more of the four definitions described above.
:::
